---
title: "Market Values in the Big5 European football leagues"
subtitle: "An Exploratory Data Analysis with Transfermarkt data"
author: "David Desquens"
date: '`r Sys.Date()`'
format: 
  html:
    code-fold: true
    code-tools: true
    df-print: paged
    toc: true
    number_sections: false
editor: visual
---

#### Let's start exploring our dataset

```{r}
#| label: libraries-load
#| include: false

library(tidyverse)
library(ggthemes)
library(ggridges)
library(viridis)

 dat <- readRDS(gzcon(url("https://github.com/JaseZiv/worldfootballR_data/raw/master/data/tm_player_vals/big5_player_vals.rds")))
str(dat)

#calculate max year in the dataset
max.season <- max(dat$season_start_year)
```

First of all, data prep, we apply some operations of data cleansing & wrangling

```{r}
#| label: data-prep

#we only want a record per player per year, we will only keep the last record of the year, that is the one where the player joined a squad later, the one which has the greatest date_joined, in case of resulting in multiple records, then the biggest contract_expiry will take precedence
dat.prep <- dat %>% group_by(player_name,season_start_year) %>% mutate(last.record.season = row_number()) %>% arrange(desc(date_joined),desc(contract_expiry)) %>% filter(last.record.season==1)

#get rid of players without market value
dat.prep <- dat.prep %>% filter(!is.na(player_market_value_euro))

#check how many records without market value
#summary(dat.prep$player_market_value_euro)

#check how many players have more than one record per year, are they legit records?
#check.records.players <- dat %>% count(player_name,season_start_year,sort=T) %>% filter(n > 1)
#print(check.records.players)

#let's check the first one, Adama Traoré in 2019
#print(filter(dat,player_name=='Adama Traoré') %>% arrange(season_start_year,date_joined,contract_expiry) %>% select(season_start_year,squad,date_joined,contract_expiry,player_market_value_euro))
```

How many records do we have and since which year?

```{r}
#| label: records-year
#| include: true

#barplot of records per year
barplot.records.year <- ggplot(dat.prep,aes(season_start_year)) + geom_bar() + theme_clean() + labs(x='season',y='# players')
print(barplot.records.year)
```

What's the total aggregated Market Value per season in the Big5?

```{r}
#| label: value-year
#| include: true

#col plot of market value per year
ggplot(dat.prep,aes(season_start_year,round(player_market_value_euro/1e6),1)) + geom_col(aes(fill=country)) + theme_clean() + labs(x='season',y='market value (€M)')
```

What is the market value of each position for each of the big5 in the last season `r max.season`?

```{r}
#| label: value-position-country

#use faceting to split the colplots based on countries
ggplot(dat.prep %>% filter(season_start_year==max.season),aes(player_market_value_euro/1e6,player_position,fill=country)) + geom_col() + theme_clean() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + facet_grid(. ~ country) + labs(y='position',x='market value (€M)')
```

How much could it cost Barça signing both Joãos?

```{r}
#| label: joaos

#let's plot the market values of the Joaos through seasons ("João Cancelo","João Félix")
ggplot(dat.prep %>% filter(player_name %in% c("João Cancelo","João Félix")),aes(season_start_year,player_market_value_euro/1e6)) + geom_line(aes(group=player_name,color=player_name)) + geom_point(aes(color=player_name)) + theme_calc() + labs(x='season',y='market value (€M)')
```

Are there any differences in `player_age` among the big5?

```{r}
#| label: test

#let's plot the market values of the Joaos through seasons ("João Cancelo","João Félix")
ggplot(dat.prep,aes(x=player_age,y=country, fill = after_stat(x))) + geom_density_ridges_gradient(scale=3,gradient_lwd=1.,alpha=.1) + scale_fill_gradientn(name="",colours=rev(viridis::viridis(5)),limits=c(10,45)) + theme_fivethirtyeight() + theme(legend.position = 'right', legend.direction = 'vertical') + ggtitle("Player Age distribution by Country")
```

-   France, Germany and Italy are predominated by youngsters vs. England and Spain where teams are oldish
